---
title: "Motor Trend Analysis"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r, echo=FALSE}

library(ggplot2)

```

## 1. Summary

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

1. ¡°Is an automatic or manual transmission better for MPG¡±
2. "Quantify the MPG difference between automatic and manual transmissions"

## 2. Exploratory Data Analyses

```{r}

data(mtcars)
dim(mtcars)

```

As we can see, the *mtcars* data set has `r ncol(mtcars)` variables and `r nrow(mtcars)` observations.

Based on the data description, *am*, *vs*, *cyl*, *gear*, *carb* are discrete. (Code is in the appendix part 1)

```{r, echo=FALSE}

mtcars$am <- factor(mtcars$am, levels=0:1, labels = c("Automatic", "Manual"))
mtcars$vs <- factor(mtcars$vs, levels=0:1, labels = c("Vshaped", "Straight"))
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$gear <- as.factor(mtcars$gear)
mtcars$carb <- as.factor(mtcars$carb)

```

In order to explore the relationship between a set of variables and MPG, we can draw scatterplots for each two variables. (In the appendix part 2)

## 3. Is an automatic or manual transmission better for MPG

In order to find whether there is difference between automatic and manual transmission for MPG, we can draw a boxplot to show it. (In the appendix part 3)

We can also fit a linear regression model to check it.

```{r}

fit1 <- lm(mpg ~ am, mtcars)
(fit1_summ <- summary(fit1))

```

The p-value is smaller than 0.05, so we reject the null hypothesis, which means there exists difference between automatic and manual transmissions. Actually, the coefficient of *Manual* is `r round(fit1_summ$coefficients[2, 1], 4)`, which means manual transmission has `r round(fit1_summ$coefficients[2, 1], 4)` MPG larger than automatic, thus the automatic is better than the manual.

However, we notice that the R-squared is only `r round(fit1_summ$r.squared, 4)`, which means only `r round(fit1_summ$r.squared, 4)*100`% variance in MPG can be explained by transmission types. Therefore, we will consider to add more variables into the linear model.

## 4. Quantify the MPG difference between automatic and manual transmissions

First of all, fit a full model. Then we will do a stepwise regression for it, where the choice of predictor is carried out automatically by comparing AIC (the smaller AIC, the better model). Details are in the appendix part 4.

```{r}

fit2 <- lm(mpg ~ ., mtcars)
fit3 <- step(fit2, trace=0)
(fit3_summ <- summary(fit3))

```

Finally, we get a best model with four predictors: number of cylinders, gross horsepower, weight, and transmission. The R-squared shows there is `r round(fit3_summ$r.squared, 4)*100`% variance can be explained, which is good enough.

The diagnostic plots also show this model is good. (In the appendix part 5)

If other variables are fixed, then the MPG will decrease 0.03211 as gross horsepower increases one unit, the MPG will decrease 2.49683 as weight increases 1000 lbs, the manual transmission will lead the MPG has 1.89921 higher than automatic transmission. As for number of cylinders, compring to 4 cylinders, 6 cylinders will decrease 3.03134 MPG, and 8 cylinders will decrease 2.16368 MPG.

\newpage

## Appendix

### 1. Code for transform numeric variables to factor variables

```{r, echo=FALSE}

data(mtcars)

```

```{r}

mtcars$am <- factor(mtcars$am, levels=0:1, labels = c("Automatic", "Manual"))
mtcars$vs <- factor(mtcars$vs, levels=0:1, labels = c("Vshaped", "Straight"))
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$gear <- as.factor(mtcars$gear)
mtcars$carb <- as.factor(mtcars$carb)

```

### 2. Scatterplots for each two variables.

```{r}

pairs(mpg ~ ., data = mtcars, main="Relationships Between Variables and MPG")

```

### 3. Boxplot for the difference between automatic and manual transmission for MPG.

```{r, fig.height=3, fig.width=5.5}

ggplot(mtcars, aes(x=factor(am), y=mpg, fill=factor(am))) + geom_boxplot() +
    labs(x="Transmissions", y="Miles/(US) Gallon",
         title="Miles Per (US) Gallon for Different Transmissions") +
    guides(fill=guide_legend(title="Transmissions"))

```

As the boxplot shows, the automatic transmission is significant better than manual transmission for MPG. 

### 4. Details for stepwise regression.

```{r}

step(fit2)

```

### 5. Diagnostic plots for the final model.

```{r}

par(mfrow = c(2, 2))
plot(fit3)

```